---
title: "Mito 03 — Differential Expression"
subtitle: "Design, contrasts, model fit, and the Π-score, on the non-imputed matrix"
author: "A_Mito_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces the primary DEP stage and shows the result of each step.
It runs the same four-step proteoDA workflow as
`03_DEP/a_non_imputed/a_script/01_run_dep.R` — describe the design, define the
comparisons, fit the model, pull the results — live and in memory. The headline
counts are computed from the live fit, then checked against the committed output.

# Start from the non-imputed matrix {#sec-load}

The test runs on the matrix that still has its missing values; `limma` handles
NAs per protein:

```{r}
#| label: load
library(proteoDA)
dal <- readRDS(here("02_Normalization", "c_data", "DAList_normalized.rds"))
dim(dal$data)
table(dal$metadata$group)                 # four groups, six each
```

# Step 1 — describe the design {#sec-design}

`add_design()` turns a formula into a model matrix. `~ 0 + group` gives each of
the four groups its own column (its own mean); `(1 | Replicate)` tells the model
that wells sharing a replicate index are paired:

```{r}
#| label: add-design
dal <- add_design(dal, "~ 0 + group + (1 | Replicate)")
colnames(dal$design)
head(dal$design)
```

The `(1 | Replicate)` block is not a column; proteoDA passes it to
`limma::duplicateCorrelation`, so the paired structure costs one correlation
parameter instead of five degrees of freedom.

# Step 2 — define the comparisons {#sec-contrasts}

The 2×2 design gives five readable contrasts:

```{r}
#| label: add-contrasts
contrasts <- c(
  "CTLvPHE       = PHE - Ctl",                        # Disease
  "CTLvMITO      = Mito - Ctl",                        # Intervention
  "PHEvPHE_MITO  = PHE_Mito - PHE",                    # Rescue
  "Interaction   = (PHE_Mito - PHE) - (Mito - Ctl)",  # Interaction
  "MITOvPHE_MITO = PHE_Mito - Mito")                  # Secondary
dal <- add_contrasts(dal, contrasts_vector = contrasts)
dal$contrasts
```

Read each column downward to see which group means are differenced — `PHEvPHE_MITO`
(the rescue) is `PHE_Mito` minus `PHE`.

# Step 3 — fit the model {#sec-fit}

One call estimates the within-replicate correlation, fits the model with it, and
applies robust empirical-Bayes moderation:

```{r}
#| label: fit
dal <- fit_limma_model(dal)
dal$eBayes_fit$correlation     # consensus within-replicate correlation
```

The correlation is small here — the replicates are well balanced, so the block
adds little, but it is still correct to include.

# Step 4 — pull the results {#sec-extract}

`extract_DA_results()` builds a tidy table per contrast into `$results`:

```{r}
#| label: extract
dal <- extract_DA_results(dal, pval_thresh = 0.10, lfc_thresh = 0, adj_method = "BH")
names(dal$results)
head(dal$results$PHEvPHE_MITO[, c("logFC", "P.Value", "adj.P.Val")])   # the rescue contrast
```

# The Π-score {#sec-pi}

The Π-score (Xiao 2014, $\Pi = \text{P.Value}^{|\text{logFC}|}$) folds effect
size into significance, so it is only small when a protein is both significant and
moves a lot. `sig_pi` tags each protein `+1` / `−1` / `0` at Π < 0.05:

```{r}
#| label: pi
library(purrr); library(tibble); library(dplyr)
res <- imap(dal$results, \(r, cn) as_tibble(r, rownames = "uniprot_id") |>
  mutate(contrast = cn,
         pi_score = P.Value ^ abs(logFC),
         sig_pi = case_when(pi_score < 0.05 & logFC > 0 ~ 1L,
                            pi_score < 0.05 & logFC < 0 ~ -1L, TRUE ~ 0L)))
res$PHEvPHE_MITO |> dplyr::count(sig_pi)
```

# Headline numbers {#sec-results}

Summary across the five contrasts, computed from the live fit:

```{r}
#| label: summary
bind_rows(res) |>
  group_by(contrast) |>
  summarise(n = n(),
            FDR_0.10 = sum(adj.P.Val < 0.10, na.rm = TRUE),
            Pi_0.05  = sum(sig_pi != 0, na.rm = TRUE), .groups = "drop")
```

The rescue contrast (`PHEvPHE_MITO`) is the strongest by far — mitochondrial
transplant substantially remodels the PHE-stressed proteome. That is the headline
biological result (`Mito_pipeline.qmd §5`).

The production script additionally runs proteoDA's native writers (shown here)
and saves `results.xlsx` + `combined_results_pi.csv` to `c_data`, with plots to
`b_reports`:

```{r}
#| label: writers
#| eval: false
write_limma_tables(dal, output_dir = here("03_DEP", "a_non_imputed", "c_data"),
                   annot_cols = c("uniprot_id", "gene", "protein", "description"), overwrite = TRUE)
write_limma_plots(dal, grouping_column = "group",
                  output_dir = here("03_DEP", "a_non_imputed", "b_reports"), overwrite = TRUE)
```

The live summary matches the committed `combined_results_pi.csv`:

```{r}
#| label: verify
library(readr)
read_csv(here("03_DEP", "a_non_imputed", "c_data", "combined_results_pi.csv"),
         show_col_types = FALSE) |>
  group_by(contrast) |>
  summarise(n = n(), FDR_0.10 = sum(adj.P.Val < 0.10, na.rm = TRUE),
            Pi_0.05 = sum(sig_pi != 0, na.rm = TRUE), .groups = "drop")
```

# Did imputation matter? {#sec-concordance}

`03_DEP/b_imputed/` re-runs the model on the two imputed matrices and correlates
fold changes against this non-imputed answer:

```{r}
#| label: concordance
read_csv(here("03_DEP", "b_imputed", "c_data", "logFC_vs_nonimputed.csv"),
         show_col_types = FALSE) |> as.data.frame()
```

Both methods track the non-imputed result closely (ρ ≈ 0.95+) — Mito's low
missingness means the imputation choice barely moves the answer, so the
non-imputed result is the one reported.

# Recap {#sec-recap}

1. `add_design()` — a column per group, plus the Replicate block.
2. `add_contrasts()` — name-equals-difference comparisons (the 2×2 questions).
3. `fit_limma_model()` — one consensus correlation + robust eBayes.
4. `extract_DA_results()` — tidy `$results`, one table per contrast.
5. The Π-score for effect-size-aware calls, then proteoDA's native writers.

The test runs on the non-imputed matrix; `b_imputed/` is a concordance check. The
figures in `04_Figures/` build on these contrasts (`Mito_pipeline.qmd §6`).
