---
title: "Mito 02 — Normalization & Imputation"
subtitle: "Cyclic-loess normalization, and when to fill missing values"
author: "A_Mito_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces stage 02 of the Mito pipeline and shows the result of
each step. It runs the same logic as `02_Normalization/a_script/01_run_normalization.R`
(plus the three imputation arms in `02_Normalization/imputation/a_script/`), in
memory, writing no files.

# Load the filtered `DAList` {#sec-load}

Stage 01 hands off a filtered, un-normalized object that still carries its
(small) missingness:

```{r}
#| label: load
library(proteoDA)
dal <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
miss_before <- round(100 * mean(is.na(dal$data)), 1)
dim(dal$data)
miss_before # small but non-zero missingness, not yet imputed
dal$tags
```

# Normalize with cyclic loess {#sec-normalize}

Samples load onto the instrument slightly differently, so whole samples sit
higher or lower for technical reasons. `normalize_data(norm_method = "cycloess")`
evens out each sample's median log-intensity in one native call:

```{r}
#| label: normalize
before <- apply(log2(as.matrix(dal$data)), 2, median, na.rm = TRUE) # filtered data still linear
dal <- normalize_data(dal, norm_method = "cycloess")
after <- apply(as.matrix(dal$data), 2, median, na.rm = TRUE) # cycloess already returns log2

rbind(before = summary(before), after = summary(after)) # spread collapses to a tight band
dal$tags # new 'normalized' + 'norm_method' tags
```

Normalization **rescales** but does not fill anything in — the missingness is
unchanged:

```{r}
#| label: missing-unchanged
c(
  before_pct = miss_before,
  after_pct = round(100 * mean(is.na(dal$data)), 1)
)
```

The script also writes proteoDA's before/after audit PDFs to `b_reports/`:

```{r}
#| label: qc-reports
#| eval: false
write_norm_report(dal,
  grouping_column = "group", output_dir = report_dir,
  filename = "norm_comparison.pdf"
)
write_qc_report(dal,
  color_column = "group", output_dir = report_dir,
  filename = "qc_normalized.pdf"
)
```

This normalized, **non-imputed** object is the primary hand-off to stage 03 —
`limma` handles missing values per protein.

# Imputation for methods that need a complete matrix {#sec-impute}

PCA, WGCNA, the MitoCarta set scores, and the rotation tests need a matrix with
no gaps. Three independent methods are kept so the choice can be checked: the
two mechanism-aware ones shown below, plus a missForest random-forest arm
(`c_missforest.R`) carried for comparison.

## imp4p (canonical imputation arm) {#sec-imp4p}

`imp4p` is mechanism-aware: it estimates the per-protein mix of missing-at-random
vs too-low-to-detect values and imputes each accordingly. It is stochastic, so
the seed is fixed:

```{r}
#| label: imp4p
#| cache: true
library(imp4p)
set.seed(42)
mat <- as.matrix(dal$data)
cond <- factor(dal$metadata$group[match(colnames(mat), dal$metadata$Col_ID)])
imp4p_mat <- as.matrix(impute.mi(
  tab = mat, conditions = cond,
  methodMCAR = "mle", methodMNAR = "igcda", progress.bar = FALSE
))
dimnames(imp4p_mat) <- dimnames(mat)

c(
  dim_proteins = nrow(imp4p_mat), dim_samples = ncol(imp4p_mat),
  na_remaining = sum(is.na(imp4p_mat))
) # 0 remaining — complete matrix
```

## MsCoreUtils hybrid (second opinion) {#sec-mscoreutils}

The second method classifies each protein MAR vs not via
`imputeLCMD::model.Selector()`, then routes MAR features to kNN and MNAR features
to a left-censored QRILC draw:

```{r}
#| label: mscoreutils
library(MsCoreUtils)
library(imputeLCMD)
set.seed(42)
randna <- as.logical(model.Selector(mat)[[1]]) # TRUE = MAR feature
ms_mat <- impute_matrix(mat, method = "mixed", randna = randna, mar = "knn", mnar = "QRILC")

c(
  MAR_features = sum(randna), MNAR_features = sum(!randna),
  na_remaining = sum(is.na(ms_mat))
)
```

Mito has little missingness to begin with, so the two methods end up very close
(stage 03 quantifies it). They only differ where a value was actually missing:

```{r}
#| label: compare
was_missing <- is.na(mat)
c(
  entries_imputed = sum(was_missing),
  pct_of_matrix = round(100 * mean(was_missing), 1)
)
```

# Where stage 02 stands {#sec-recap}

The normalized matrix is saved to `02_Normalization/c_data/`; the three imputation
arms write to `02_Normalization/imputation/c_data/`:

| File | Location | Missing values? | Used by |
|---|---|---|---|
| `DAList_normalized.rds` | `c_data/` | kept | the DE test (stage 03) — **primary** |
| `DAList_imputed_imp4p.rds` | `imputation/c_data/` | filled | figures, WGCNA, rotation (canonical imputation arm) |
| `DAList_imputed_mscoreutils.rds` | `imputation/c_data/` | filled | the robustness comparison |
| `DAList_imputed_missforest.rds` | `imputation/c_data/` | filled | the random-forest comparison |

Normalization levels the samples but never invents data; imputation is applied
only for methods that cannot run on an incomplete matrix, and never before the DE
test.
